---
title: "Final Project"
author: "Coco Donovan"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

## Abstract:

The California Racial and Identity Profiling Act of 2015 (AB 953) mandated, 
among other items: law enforcement is prohibited from profiling based on race/
ethnicity or identity and that law enforcement be required to submit data on 
vehicle/pedestrian stops and citizen complaints alleging racial and identity 
profiling.

The primary dataset I analyzed comes from the reported Police Stop data from the
Long Beach Police Department between Jan 1st, 2019, and Dec 31st, 2021. In this 
analysis, I intend to focus specifically on the differences in policing people 
of different ethnicities as measured by stop duration and the number of actions 
taken by the Long Beach Police Department during the stop.

### Loading Necessary Packages

```{r}
library(dplyr)
library(tidyverse)
library(openintro)
library(infer)
library(ggplot2)
```

### Data Preparation

```{r}
lb_police_stops <- read.csv('lbpd-ripa-data-annual.csv')

# head(lb_police_stops)

# colnames(lb_police_stops)
```

Limitations: When comparing police "perceived race" with official data on ethnicity for 
Long Beach the comparison it is clear that there is less nuance to the police
"race" data. I understand identity is far more complex than the way I have simplified
it for the purpose of analysis, and for the purpose of this less extensive analysis
my simplified classifications will be used to present the data.

__________________________________________________________________________

### Population of Long Beach, CA by Ethnicity

```{r}
lb_population <- read.csv('Race and Ethnicity.csv')

lb_population <- lb_population %>%
  filter(Year == 2020)

# lb_population

lb_population_eth_percentage <- lb_population %>%
  mutate(eth_true = ifelse(Ethnicity == 'Hispanic or Latino', 'Hispanic', Race)) %>%
  group_by(eth_true) %>%
  summarize(percentage = round(100 * sum(share),2)) %>%
  arrange(desc(percentage))

knitr::kable(lb_population_eth_percentage)
```

### Visualization

```{r}
ggplot(data = lb_population_eth_percentage, mapping = aes(y = reorder(eth_true, percentage), x = percentage, fill = eth_true)) +
  geom_bar(stat = 'identity') + 
  labs(y = 'Ethnicity', x = 'Percent of Population', fill = 'Ethnicity', title = "Percent of Population of Long Beach, CA by Ethnicity") + 
  theme(legend.position = "none")
```

_________________________________________________________________________

### Percentages by Ethnicity Breakdown of all Stops

Note: maybe pull the demographics of long beach and compare for proportionality
I know it's two data sources, but I'll live

```{r}
Percent_by_Eth <- lb_police_stops %>%
  select("Perceived.Race.Simplified") %>%
  group_by(Perceived.Race.Simplified) %>%
  summarise(Percent_of_Stops = 100 * n()/nrow(.), total_stops = n()) %>%
  arrange(desc(Percent_of_Stops))

knitr::kable(Percent_by_Eth)
```

### Visualization

```{r}
ggplot(data = Percent_by_Eth, mapping = aes(y = reorder(Perceived.Race.Simplified, Percent_of_Stops), x = Percent_of_Stops, fill = Perceived.Race.Simplified)) +
  geom_bar(stat = 'identity') + 
  labs(y = 'Perceived Ethnicity', x = 'Percent of Total Stops', fill = 'Perceived Ethnicity', title = 'Percent of Total Stops by Ethnicity') +
  theme(legend.position = "none")
```

_________________________________________________________________________

### Ethnicity vs Duration of Stop

```{r}
# colnames(lb_police_stops)

Stop_duration_by_Eth <- lb_police_stops %>%
  select("Perceived.Race.Simplified", "Stop.Duration") %>%
  group_by(Perceived.Race.Simplified) %>%
  summarise(avg_duration = mean(Stop.Duration)) %>%
  arrange(desc(avg_duration))

knitr::kable(Stop_duration_by_Eth)
```

### Visualization

```{r}
ggplot(data = Stop_duration_by_Eth, mapping = aes(y = reorder(Perceived.Race.Simplified, avg_duration), x = avg_duration, fill = Perceived.Race.Simplified)) +
  geom_bar(stat = 'identity') + 
  labs(y = 'Perceived Ethnicity', x = 'Stop Duration', fill = 'Perceived Ethnicity', title = "Average Stop Duration by Ethnicity") +
  theme(legend.position = "none")
```

### 95% Confidence Intervals of Stop Duration by Ethnicity

```{r}
summary_stops_by_eth <- lb_police_stops %>% 
  group_by(Perceived.Race.Simplified) %>% 
  summarise(sd_time = sd(Stop.Duration, na.rm = TRUE), mean_time = mean(Stop.Duration, na.rm = TRUE), count_stops = sum(table(Stop.Duration)))

summary_stops_by_eth <- summary_stops_by_eth %>%
  mutate(lower_ci = mean_time - 1.96 * (sd_time / sqrt(count_stops)), upper_ci = mean_time + 1.96 * (sd_time / sqrt(count_stops)))

arranged_summary_stops <- summary_stops_by_eth %>%
  select(Perceived.Race.Simplified, lower_ci, upper_ci, count_stops) %>%
  arrange(desc(lower_ci))

knitr::kable(arranged_summary_stops)
```

### Visualization

```{r}
ggplot(arranged_summary_stops) +
  geom_segment( aes(x=reorder(Perceived.Race.Simplified, lower_ci), xend=Perceived.Race.Simplified, y=lower_ci, yend=upper_ci), color="black") +
  geom_point( aes(x=Perceived.Race.Simplified, y=lower_ci), size=3 ) +
  geom_point( aes(x=Perceived.Race.Simplified, y=upper_ci), size=3 ) +
  coord_flip()+
  theme(
    legend.position = "none",
  ) +
  xlab("") +
  ylab("Stop Duration (Minutes)") +
  labs(title= "95% Confidence Interval by Ethnicity for Stop Duration (Minutes)")
```

__________________________________________________________________________

### Ethnicty vs Action Taken During Stop

```{r}
Action_Taken_by_Eth <- lb_police_stops %>%
  select("Perceived.Race.Simplified", "Num.Actions.Taken") %>%
  group_by(Perceived.Race.Simplified) %>%
  summarise(avg_num_of_actions = mean(Num.Actions.Taken)) %>%
  arrange(desc(avg_num_of_actions))

knitr::kable(Action_Taken_by_Eth)
```

### Given a person stopped is black, what is the likelihood action is taken? (Sampling)

```{r}
black_and_action_taken <- lb_police_stops %>%
  filter(Perceived.Race.Simplified == 'Black/African American') %>%
  mutate(action_taken = ifelse(Actions.Taken != 'None', "yes", "no"))

black_and_action_taken %>%
  filter(!is.na(action_taken)) %>%
  specify(response = action_taken, success = "yes") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "prop") %>%
  get_ci(level = 0.95)
```

### Given and person is Hispanic or Latino, what is the likelihood a police will take action

```{r}
hispanic_and_action_taken <- lb_police_stops %>%
  filter(Perceived.Race.Simplified == "Hispanic/Latino/a") %>%
  mutate(action_taken = ifelse(Actions.Taken != 'None', "yes", "no"))

hispanic_and_action_taken %>%
  filter(!is.na(action_taken)) %>%
  specify(response = action_taken, success = "yes") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "prop") %>%
  get_ci(level = 0.95)
```

### Given a person is white, what is the likelihood of police action?

```{r}
white_and_action_taken <- lb_police_stops %>%
  filter(Perceived.Race.Simplified == 'White') %>%
  mutate(action_taken = ifelse(Actions.Taken != 'None', "yes", "no"))

white_and_action_taken %>%
  filter(!is.na(action_taken)) %>%
  specify(response = action_taken, success = "yes") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "prop") %>%
  get_ci(level = 0.95)
```

### Given a person is Asian, what is the likelihood of police action?

```{r}
Asian_and_action_taken <- lb_police_stops %>%
  filter(Perceived.Race.Simplified == 'Asian') %>%
  mutate(action_taken = ifelse(Actions.Taken != 'None', "yes", "no"))

Asian_and_action_taken %>%
  filter(!is.na(action_taken)) %>%
  specify(response = action_taken, success = "yes") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "prop") %>%
  get_ci(level = 0.95)
```

__________________________________________________________________________

### Does more time with the police Mean more actions taken?

```{r}
lb_police_stops %>%
  summarise(cor(Stop.Duration, Num.Actions.Taken, use = "complete.obs"))

lb_police_stops %>%
  filter(Perceived.Race.Simplified == "Black/African American") %>%
    summarise(cor(Stop.Duration, Num.Actions.Taken, use = "complete.obs"))
```

__________________________________________________________________________

## Conclusion:

In conclusion, we can see from the data that the Black Community in Long Beach 
experiences disproportionately long Police Stops when compared to people of other
races (with exception of the ethnicities that made up smaller portions of the 
overall population i.e. Native American and Pacific Islander populations). The 
Black community also experienced a disproportionate amount of police action, when 
compared to the ethnicities that made up larger portions of the population.

When I quickly checked to see if there was correlation between the stop duration
and the number of actions taken, there was only weak correlation at best. 

Though what is clear is that the Black community of long beach faces statistically 
significant unfair treatment at the hands of the Long Beach Police Department 
during Police Stops.

### Additional Analysis Possibilities:

This is something I did not have time for, but a cool addition to this sort of 
analysis would be to classify actions by Violent vs Non Violent (Use of Force vs
Non-Use-of-Force). I would have to do some reading to see which actions are
classified by department definition as use of force and then repeat this sort of
analysis with the a binary representing whether or not an encounter involved the 
use of force and the counted number of uses of force.
